---
title: "Week 1. Introduction to R"
author: "Yuriy Davydenko"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    includes:
      in_header: "favicon.html" 
---

*** 

<br>

#### Installing R & RStudio

__Step 1:__ Go to the  *R Project for Statistical Computing*  at [www.r-project.org](www.r-project.org), download and install R 

__Step 2:__ Go to  RStudio  at [https://rstudio.com/products/rstudio/](https://rstudio.com/products/rstudio/),  download and install RStudio Desktop (Open Source Edition)

<br>

__*ALTERNATIVELY:*__

Use  RStudio Cloud  at https://rstudio.cloud

* Create personal account
* In every class, use the link that I provide to access the class material

<br>

### R Primer

***

#### RStudio Cloud

Use the following [link](https://rstudio.cloud/project/802723) to access today's workshop on RStudio Cloud.


<br>

#### RStudio Interface

* Console, 
* Source, 
* Environment, 
* Viewer 

<br>

#### Setting work directory
```{r}
getwd()
setwd(".")
```

Commenting and executing commands:
```{r}
# This is a comment
# There is no capacity for multi-line comments although you can highlight multiple lines by pressing Shift+Ctl+C
# Execute command with Ctrl+Enter
```

<br>

#### Data Structures
Four primary data structures in R:

* Vectors
* Matrices
* Data Frames
* Lists

Vectors: 

* Numeric
* Integer
* Character
* Factor
* Logical


#### Scalars

```{r}
2 + 2   # Basic Math
sqrt(4)  # Square root of four
```

#### Variables & assignment operator

```{r}
x <- 1      # <- "gets" operator
x           # Displays the values in x
y <- 2
y
z <- x+y
z
```

#### Other Operators
```{r}
1:10    # ":" produces integers
c(1, 2, 3, 4)     # produces floating-point numbers
1==1    # Test for equality
1!=1    # Test for inequality
# <, >, <=, >=   # more tests
```


```{r}
1:10    # Prints numbers 1 to 10
print("Hello World!")
```



#### Vectors
Vectors - the R workhorse, a container for values
Let's make a simple dataset (vector):
```{r}
scores1 <- c(1, 2, 3, 4, 5)   # Puts the numbers 1-5 in the variable x
scores2 <- 1:5                # Puts the numbers 1-5 in the variable x
scores3 <- c(scores1, scores2, 1, 2, 3, 5)
seq(from=1, to=5, by=1)
rep(1, times=5) 
```

#### Vector Math
```{r}
x * 2  # Multiplies each element in x by 2
x + y  # Adds corresponding elements in x and y
```

#### Applying functions to vectors:
```{r}
max(x)
min(scores3)
length(x)
```

#### Indexing/subsetting vectors:
```{r}
x[1]  # chooses 1st element of the vector x
x[2:3]

v <- c(1, 15, 2020); names(v) <- c("Month", "Day", "Year")
v["Year"]	# If vector elements have names - select by name

```

#### Negative subscripts mean that we want to exclude the given elements in our output.
```{r}
x[-1]    # exclude element 1
x[-1:-2] # exclude elements 1 through 2
```


#### Logical operations
```{r}
v <- 1:10
v>5
```


#### Filtering
```{r}
v <- seq(from = 1, to = 20, by = 2)
v[v>5]
subset(v,v > 5)  # NA values are removed when using subset()
which(v>5)       # find the positions within v at which the condition occurs
```


#### Transforming
```{r}
v <- 1:10
v
v.dummy <- ifelse(v>5, 1,0)  # creating a binary variable 
```


#### Characters
```{r}
a <- c(1,2,3)
b <- "1,2,3"
c <- paste("abc","de","f") # concatenate the strings
d <- strsplit(c," ") # split the string according to blanks

length(a); length(b)
```

#### Matrices
A matrix - a rectangular array of numbers
```{r}
m <- matrix(1:4, nrow = 2)    # Create a 2x2 matrix
m <- matrix(1:4, ncol = 2)
m <- rbind(c(1,4),c(2,2))
```

Indexing/subsetting a matrix:
```{r}
m[1,]      # select firs row
m[,2]      # select second column
m[2,1]     # select the element in the second row, first column
```

#### Lists
An R list is a container for values, but its contents can be items of different data types.
```{r}
l <- list(test_scores=25, student_names="xyz")
l
```

The expression `l$name1` refers to the name1 component in the list l:
```{r}
l$name1
str(l)
```

#### Data Frames
A data frame in R is a list, each element of which is a vector of the same lenght
```{r}

df <- data.frame(Month = c("Jan", "Feb", "March"), Day = c(15, 20, 22), Year = rep(2020, times = 3))
df
```

#### NA and NULL Values 
```{r}
v <- c(1,NA,3,4,5)
max(v)
max(v,na.rm=T)

v <- NA
c(v, 1)

v0 <- NULL
c(v0, 1)   # NULL values really are counted as nonexistent
```


#### Checking Data Types

```{r}
a <- 1:10; b <- "1 to 10"

typeof(a)
typeof(b)
mode(a)
mode(b)
class(a)
class(b)

df <- data.frame(Month = c("Jan", "Feb", "March"), Day = c(15, 20, 22), Year = rep(2020, times = 3))
typeof(df)
mode(df)
class(df)
```

#### Clean Up
```{r}
rm(x)  # Remove an object from workspace
rm(a, b)  # Remove more than one
rm(list = ls())  # Clear entire workspace
```


#### R-Style Guide
```{r}
browseURL("https://jef.works/R-style-guide/")
```

#### Help
```{r}
help(functionname) 
help(mean)
?mean
example(max) 
?"<"
```


####R Packages

See existing packages online:
```{r}
# browseURL("https://cran.r-project.org/web/packages/available_packages_by_name.html")
```

```{r}
# library()  # List of installed packages
# search()   # Shows packages that are currently loaded
```

Installing & loading packages:
```{r}
# install.packages("summarytools")  # Download and install package in R
# library("ggplot2")  # Load a package into R
# require("ggplot2")  # Load a package into R
```

```{r}

# library(help = "ggplot2")  # Brings up documentation on a package
# vignette(package = "grid")  # Brings up list of vignettes (examples) in editor window
# browseVignettes(package = "grid")  # Open web page with hyperlinks for vignette PDFs etc.

# update.packages()  # Checks for updates

# detach("package:ggplot2", unload = TRUE)  # detach package

# remove.packages("psytabs")   # Deletes it

```

#### Built-in Datasets

See the available datasets online:
```{r}
browseURL("http://stat.ethz.ch/R-manual/R-devel/library/datasets/html/00Index.html")
```

```{r}

# library(help = "datasets")     # See the list of available datasets
# data()                         # See the list of available datasets
# try(data(package = "rpart") )  # list the data sets in the rpart package
# ?datasets::cars                # Information about a particular dataset in the package

# library(datasets)  # Load the library with the built-in datasets
# require(datasets)  # Load the library with the built-in datasets

# data(cars)   # Load dataset cars
# str(cars)    # See the structure of the dataset cars

# data(Cars93, package="MASS") 
```

#### Import Data
```{r}
# txt.file <- read.table("Data/file.txt", header = TRUE)           
# txt.file <- read.table("Data/file.txt", header = TRUE, sep = "\t")
# csv.file <- read.csv("Data/file.csv", header = TRUE)
```

#### Saving Data
```{r}
# save(myData, file = "myData.RData")  # save the object to a file 
# load(file = "myData.RData")          # load previously saved data from .RData file
```

<br>

*** 

<br>

Download this source file:

* [R Script](docs/01_Intro.R)
* [R Markdown](docs/01_Intro.Rmd)











